Challenges and Solutions of Developing and Implementing a Novel Desktop-as-a-Service

### 2024-04-29 paper.tex
Condense chapters performance and stability

DaaS has been a well-known service category since the emergence of cloud computing. Still, it got much less attention in research and literature
The challenges and obstacles when developing DaaS solutions, rather than infrastructure (IaaS) and platform services (PaaS), have seldom been discussed in the literature.

Celesti et al.~\cite{celesti2016improving} implemented in 2016 a DaaS using the IaaS solution OpenStack~\cite{sefraoui2012openstack} and analyzed the characteristics and performance aspects of using noVNC, which is a client for the protocol Virtual Network Computing (VNC) that is implemented in the Hypertext Markup Language (HTML) and JavaScript, the protocol SPICE (Simple Protocol for Independent Computing Environments), and the remote desktop gateway Apache Guacamole as solutions for providing access to the desktop via a browser. One focus of the paper is the redirection of the sound interface of the virtual desktop. The paper's authors conclude that Guacamole, in combination with the Remote Desktop Protocol (RDP), is the best solution. The authors evaluated the lag time for remote audio playback in an Internet scenario. by measuring the time between the automated initiation of the live streaming using the media player software VLC and the arrival of the music data by analyzing network traffic via monitoring software Wireshark. When using Guacamole combined with the protocol RDP, the average lag was around 750\,ms. When using Guacamole combined with the protocol VNC, the average lag was around 1750\,ms. The paper also includes a short evaluation of the lag for video frames represented by background color changes of a Java application. Again, the configuration Guacamole and RDP offered the best performance video update with an average lag of around 300\,ms. The configuration of Guacamole and VNC caused an average lag of around 500\,ms.

@@

To describe the mission-critical performance aspects of our system,
we initially give a short overview of potential challenges,
describe bottlenecks as well as other problem areas.
We do that concerning the underlying use case model of DaaS-Systems
and then follow with a more thorough description
of our own specific design decisions and particular solutions.

From our perspective, performance, in general can be seen as the most crucial system property
which influences a majority of all other other relevant properties,
such as user experience, system responsiveness, and overall efficiency.
As a result, we had to address multifaceted
and potentially cross-concerning challenges
on several distinct layers within our architecture.

From the core infrastructure to backend services
and user-facing elements, each component individually
has to achieve specific minimum requirements in its domain
and to contribute to the overall performance goals of the whole system.
Hence, we outline key limiting factors that impede performance,
along with our experiences and tailored solutions
aimed at mitigating these challenges.
From our point of view, the three most critical performance aspects
for Daas-Systems are:
\begin{itemize}
	\item Infrastructure and Platform
	\item Network Latency and Bandwidth
	\item Internal Processing and Communication
\end{itemize}
As there is no generalized measurement system available,
we ordered them by their magnitude of impact
and based them to the best of our knowledge
on the obtained experiences during our research project.
Since each of them has their own characteristics,
specific behavior and requirements, implying a potential impact
on the overall system performance and resource utilization,
we describe them individually and in more detail
in the following separate subsections.

\subsubsection{Infrastructure and Platform}
In general, it can be said that the chosen infrastructure, platform, and utilized software
may have the most influential impact on CPU and resource consumption
which has a dramatic impact on the overall system performance.
As a DaaS combines all aspects of IaaS, PaaS, and SaaS into one system
each related performance characteristic has to be taken into account.

For Infrastructure-as-a-Service can be said
that our solution generally offers possibilities
to use either virtual machines or docker containers.
As can be seen in several scientific publications
discussing relevant performance aspects of such systems,
the obvious advantage in most categories
has a lightweight container solution compared to the virtual machine
if an identical program was tested in both environments
~\cite{felter2015updated,potdar2020performance,seo2014performance}.

On the one hand, that implies for our solution,
that a container-based approach is most likely preferable,
if that choice is available.
On the other hand, that  choice is not always available,
as not every app can be run in a container-based environment.
To maximize coverage of possible apps being able to run on our system,
we offer container-based and virtual machine-based application hosting
as a critical concept of our solution.

As a second observation in the area of PaaS, it can be added
that in most studies, there is no dramatic difference
if derivates of operating systems are compared to each other.
Nevertheless, there are derivates
which are more driven towards being lightweight and more performant
~\cite{balen2020performance,boras2020performance}.  %nix vs. nix win vs. win
% ~\cite{balen2020performance}. %win in vm
In cases where a Windows virtual machine can be replaced with a natively operating wine platform,
additional performance potentials can be leveraged
~\cite{huang2012performance}.  % win inside vm
In contrast to that, if the operating systems of different types are compared to each other,
there are more obvious results
especially when taking into account implementations of low-level operating system functionality,
such as interrupt handling, memory allocation, driver-specific implementations
or other more subtle requirements such as boot time
or the time needed for maintenance or system updates
~\cite{sergeev2022docker,sulaiman2021comparison}. % docker, win vs nix
% ~\cite{sulaiman2021comparison}. % office apps win vs nix

As a result of our research, we aim to support
any Windows or Linux-based operating system in virtual machines
but can only support Linux and Wine in containers
This is because only terminal-based Windows containers are supported by Microsoft.

\subsubsection{Network Latency and Bandwidth}
Another essential aspect when providing web-based access to DaaS apps
is that also network latency and bandwidth play a significant role
which contributes to a large degree to the responsiveness of the system
directly experienced by the user.
As the user utilizes its application through a web browser
we aim to support standardized frame-buffer protocols
to enable user interaction and access to external devices such as printers or USB sticks.

Within such protocols and in cases where video content is transmitted,
the provided network bandwidth might be the main bottleneck
especially when it comes to high-resolution video streams.
In cases where audio, mouse movement, or keyboard strokes are transmitted,
the network latency plays a more critical role.
Therefore, the experienced responsibility and usability
crucially depends on the hosted application.

In general, such application-specific requirements can, to a certain degree, be controlled
by protocol-inherent tuning parameters and utilized compression algorithms.
Therefore, our solution aims to maximize coverage of potential use cases
by offering several solutions and a set of appropriate tuning options
which intentionally forces the user to decide
which solution fits best for him.

For that purpose, we utilize the desktop gateway implementation Guacamole,
which offers support for the most utilized frame-buffer protocols, VNC and RDP
as well as all relevant tuning options.
Guacamole not only implements out-of-the-box support for the protocols itself
but also adds the ability to integrate audio support for VNC or other external devices
as well as a JavaScript-based web client.

Nevertheless, to integrate guacamole into the system,
there are still things to consider that are not handled by guacamole
but would be assumed from a full-fledged DaaS solution.

First and foremost, how authentication is handled must be considered.
On one hand, from a DaaS system, a user would expect the system to work
without forcing him to authenticate himself
to multiple authentication systems and only once per session.
On the other hand, crucial information such as user passwords
might, in general, better be kept secret from the system
so that users can not manipulate the provided system in an unintended way.
A potential problem arises if it is considered
that, by default, RDP requires a valid user session to the system.

For that reason, guacamole was integrated with a proxied approach using proxied websockets.
This enables keeping the credentials of the generated system secret to the host system
while still providing authorized but passwordless access to the hosted application.

Further, the proxy mechanism enables the support of dynamic desktop resizing
which is a default guacamole feature for RDP,
but which is not available in VNC through guacamole.
By intercepting WebSocket opcodes and listening for size requests,
the requested resolution can then be enforced
by directly interacting with the related container or virtual machine.

Both changes to the default guacamole use case are mandatory
to comply with the requirements for the underlying DaaS design
but both also impact the overall system performance and responsiveness in a negative way.
From a network perspective, the integration as a proxy adds network hop
and leads to a noticeable delay in network communication.
The protocol-level evaluation of opcodes additionally enhances that delay, and perceived responsiveness might suffer if resize requests take more time than expected.
Overall, the introduced overhead is still acceptable
and enables platform-independent and seamless access for all defined DaaS use cases.

\subsubsection{Internal processing and Communication}
Another critical aspect directly related to performance and responsiveness
is the approach in which data is internally processed, stored, and communicated
to all involved components and instances.
This includes the metadata being collected and processed by essential components
but also implies raw physical data being directly exchanged with object instances.

To provide a satisfying user experience in a DaaS system
specific components in a host system are required.
This includes access to all backend services
being used to create containers and virtual machines,
a local or remote database, a local or remote authentication service,
local and distributed filesystems, as well as
virtual or physical network devices.
Further, a web server is integrated to provide all relevant API endpoints
for the frontend communication
as well as all WebSocket endpoints for viewer communication.

Generally spoken, all of these components might become a potential bottleneck
when being utilized with high throughputs
or large amounts of concurrent applications demanding low latencies.
This is especially the case if being used in a sequential manner
or being served from the same hardware.
If delays add up, the total time needed for a request to be fully processed can increase drastically.

On an architectural level, all relevant components are, therefore designed
for being distributed to different hosting systems
if the requirements of a particular user scenario suggest it.
Nevertheless, it might still be advisable
to keep specific components directly on the host system
to ensure low latency component access.

For all components with real-time and low-latency latency usecases, it is generally advisable
to keep them as close to the core hosting system as possible.
This includes all data and communication
relevant to audio, video, mouse, and keyboard transmission,
but also the backend service pertinent to host containers and virtual machines.
That might be in the best case to host them natively on the host itself
which enables virtual network and hardware access.
In the worst case, that might be at least connected to the same physical network
to reduce network latencies as much as possible.

For all other components of the overarching nature
such as database access, authentication provider, and filesystems
it is still advisable to keep a close distance from the host
but it also highly depends on the total amounts of users
and the specifics of their use cases.
On small scales, it is still advisable to keep such components on the same system
but on larger scales, it is preferable to externalize such functionality to dedicated systems.

\subsection{Stability}
% TODO: Johannes (1)
% TODO: Christian (2)
% TODO: Johannes: Beschreibung zur Position der Komponenten, Möglichkeiten der Parallelisierung und Einschränkungen
%
% Was ist hier relevant?
% Verfügbarkeit des Service und der Benutzerdaten
% Redundante Komponenten (Hardware und Software)
% Welche Komponenten kann man sinnvoll redundant vorhalten?
% Was sind die Vor- und Nachteile?
% Auch Kosten?
% Wo sehen wir aktuell in unserer Architektur Risiken für Dienstausfall und Datenverlust?
% Features von Proxmox
% Ceph-Dateisystem
% Datenbank SQlite aktuell. Perspektivisch MariaDB
% Guacamole läuft 1x direkt auf dem Host aktuell. Läuft theoretisch auf jedem Host. Proxmox kann sich darum kümmern?!
% Docker haben/nicht haben / Vor- und Nachteile
% Hier auch auf die Skalierbarkeit beziehen
% Structure:
%	- Stability
%		- Redundancy
%		- Scalability
%	- Backend-services
%		- Database
%		- Authentication
%		- Distributed Filesystem
%		- Container/VM (Proxmox, Kubernetes?) 
%		- Instance error => Recreate
%	- Considerations:
%		- Divide and conqueror
%		- Requirement: Distributable components
%	- Risks:
%		- Filesystem error
%		- Database error
%		- Authentication error
%		- Infrastructure error (Container/VM)
%	- Benefits/Drawbacks:
%		- Distribution of components is distribution of load => Performance
%		- Eliminates bottlenecks
%		- Errors affect only errornous component

Stability in DaaS applications is of significant importance
to ensure that our solution can meet the demands of continuous
and time-critical operations across varied user scenarios.
From our point of view, any DaaS system must ensure the stability
of all system components during runtime
to maintain the ability to provide service continuity,
handle growing user loads and transparently recover from failures,
while still providing a seamless end-user experience.
Such considerations are an integral part of our architectural concept
as they directly impact the operational integrity
of the system and its hosted instances,
making them suitable for enterprise-level deployment and trust.

This definition of stability is underpinned by several architectural
and operational strategies, including redundancy of components, scalability of nodes,
and a robust backend service framework
providing means of communication between components and nodes.
These strategies, as well as their drawbacks and benefits
will be further discussed on a component level
and within the following subsections.

@@

Scalability in general, but also specifically in DaaS environments
can be approached in two fundamental ways:
Horizontal and vertical scaling.
Horizontal scaling, or scaling out, involves adding more machines or instances
to a pool to manage the increased load.
In contrast to that, vertical scaling or scaling up,
refers to adding more power (CPU, RAM)
to an existing machine~\cite{vaquero2011dynamically}.

While horizontal scaling is generally more flexible
and often preferred in cloud environments
due to its ease of integration with existing architectures,
vertical scaling remains important for applications
requiring strong single-thread performance or low latencies.

To comply with both, we defined means of configuration
such that the system itself, as well as all of its backend services
are designed to be split up into logical units
which can be distributed to dedicated systems.
Essentially, all backend services receive their dedicated configuration
and can be accessed via tuples of hostname and port.
Any component using such a backend service can obtain the relevant tuple
and then use it to fetch requests from that service.
By default and in a standalone configuration,
all backend services reside on the same native host, and relevant tuples can be derived from the host configuration.
In a distributed context, a selection of these services
can be distributed to a dedicated and remote system
to comply with horizontal scaling principles.
The relevant tuple can then be derived from any node in the system
by utilizing a common database containing the relevant meta information.

Vertical scaling of DaaS nodes is currently not planned
but administrators are free to use custom 3rd party toolchains, e.g., for GPU virtualization.
Although vertical scaling on a host basis
is beyond the scope of our research project,
it is more plausible to scale the instances
being hosted within that system and for both dimensions.

For vertical scaling of instances, appropriate parameters and API endpoints are present
to manipulate the configuration of their resource assignments,
such that arbitrary configurations are possible.

For horizontal scaling of instances
the builtin mechanisms of each backend service are automatically configured and utilized
such that instances can be arbitrarily distributed across a pool of connected nodes.
In the case of virtual machines, the responsible backend service Proxmox implements
such mechanisms as part of its 'Proxmox VE High Availability (HA)' cluster.
In the case of containers, any backend service fulfilling
similar concepts can be integrated, for example, with Kubernetes.
Within such clusters, any DaaS instance is entirely agnostic to its hosting system
and must only be reachable via a known hostname or IP.

@@

Redundancy is critical for most distributed systems
but especially for DaaS architectures,
and serves as a safeguard against service interruptions and data loss
by duplicating essential components, functions, and data.
This design principle ensures that systems remain highly available and reliable,
even during component failures.
In the following, a list of components eligible for redundancy is given
which briefly describes the most relevant strategies contained in our architecture.

\begin{itemize}
\item \textbf{Frontend/Backend:} In our final version
redundancy of the frontend and backend components will offer means of configuration
such that multiple instances can be distributed to dedicated systems
and can be the target of standardized load-balancing strategies.
By doing so, the frontend or the backend may act as a gateway
between the user and the backend service containing the relevant instances.
In that manner, redundancy is achieved, and frontend or backend
can forward requests and responses
by identifying the correct communication endpoint
based on URL's or specific parts of them,
for example, by extracting hostnames
or specific identifiers being unique within the DaaS domain.
\item \textbf{Database:} The database is integrated as a backend service
and as such, completely unaware of surrounding components
and contexts in which it is used.
The database can, therefore, be distributed to a dedicated system if needed.
Generally, any modern database solution (e.g., MariaDB) implements rich features
to maintain specific data consistency strategies
such as data replication, clustering,
dedicated failover mechanisms or also RAID control.
Therefore, redundancies in our database system are currently not planned,
but can easily be applied by users with such needs
and by using database-specific features and a modified database adapter.
\item \textbf{Authentication: } Authentication within our architecture
is also implemented as a backend service
and can, therefore, be distributed to any dedicated system,
can be a target of standardized load-balancing mechanisms and strategies
and is as well fully agnostic of surrounding components
and calling contexts.
The authentication system maintains
its credential registry in a separate database,
a customizable set of user and group privileges,
and implements the well-defined OAuth 2.0 standard
to allow remote authentication.
A customizable amount of redundant systems might be configured in our final version.
	      % TODO: Discuss: Is that true? Do we want to make Darios authenticator redundant?
\item \textbf{Distributed Filesystem:} Within our architecture
we use the distributed filesystem Ceph, which is a distributed storage system
providing object, block, and file storage. In Ceph, redundancy is achieved
through its self-healing and self-managing capabilities utilizing an algorithm
called CRUSH (Controlled Replication Under Scalable Hashing).
Data availability and fault tolerance are obtained
by automatically managing replication
of data pieces across different devices.
Our designed architecture uses these capabilities
to reliably provide
file storage to host systems and all hosted DaaS instances
and provides block storage to the infrastructural backend services.
By hosting DaaS instances directly from the distributed block storage,
arbitrary redundancy scenarios might be covered
by using specific configuration parameters.
\item \textbf{Instances:} Redundancy of instances
It is not a necessary condition for DaaS systems in general, but applying such strategies under certain conditions might make sense.
By using highly available clusters within our backend services
redundancy can be achieved by using the corresponding API endpoints
previously described and by parameterizing them appropriately.
\end{itemize}

\subsubsection{Internal Communication}
After discussing the principles being used to achieve scalability and redundancy
it adheres to describe how communication flows are maintained
in such a distributed environment.

Our proposed solution is, at the core, designed
as a message-oriented architecture
which additionally aims to fulfill DaaS properties
for scalability and redundancy as previously described.

That implies fundamentally that any action being triggered,
may it either be in the frontend on the client side,
in the backend on the host side
or within any of the hosted instances
must always fulfill the condition that the required information to execute that action
must either be directly derivable from a request or a corresponding response
or otherwise, must be immediately available to the component executing that action.

We achieve that by storing all runtime information of instances in the distributed block storage,
by storing all relevant configuration files in our distributed file storage,
by utilizing a common database to persist all object metadata
and by using our decentralized OAuth component to handle authentication.

By doing so, any change being triggered by a node within the DaaS network
is immediately available to any other node in the system
which relies on that information change.
Apart from that, any additional information needed by a component
is fully contained within the related request or response.

Therefore, arbitrarily distributed configurations are possible
as long as request and response messages reach their desired target, eventually
and as long as the availability of all relevant backend services is guaranteed.

If that is the case, each distributed DaaS node can typically
just contain all time-critical components at a bare minimum
to maintain an acceptable performance throughout the whole user-facing network stack.
This would typically include the backend service hosting the relevant instances
as well as the WebSocket proxy directly connected to the guacamole daemon
and all other components may reside somewhere in the DaaS network.

>>>>

DaaS has been a well-known service category since the emergence of cloud computing. Still, it got much less attention in research and literature.
The challenges and obstacles when developing DaaS solutions, rather than infrastructure (IaaS) and platform services (PaaS), have seldom been discussed in the literature.

Celesti et al.~\cite{celesti2016improving} implemented in 2016 a DaaS using the IaaS solution OpenStack~\cite{sefraoui2012openstack} and analyzed the characteristics and performance aspects of using noVNC, which is a client for the protocol Virtual Network Computing (VNC) that is implemented in the Hypertext Markup Language (HTML) and JavaScript, the protocol SPICE (Simple Protocol for Independent Computing Environments), and the remote desktop gateway Apache Guacamole as solutions for providing access to the desktop via a browser. One focus of the paper is the redirection of the sound interface of the virtual desktop. The paper's authors conclude that Guacamole, in combination with the Remote Desktop Protocol (RDP), is the best solution. The authors evaluated the lag time for remote audio playback in an Internet scenario by measuring the time between the automated initiation of the live streaming using the media player software VLC and the arrival of the music data by analyzing network traffic via monitoring software Wireshark. When using Guacamole combined with the protocol RDP, the average lag was around 750\,ms. When using Guacamole combined with the protocol VNC, the average lag was around 1750\,ms. The paper also includes a short evaluation of the lag for video frames represented by background color changes of a Java application. Again, the configuration Guacamole and RDP offered the best performance video update with an average lag of around 300\,ms. The configuration of Guacamole and VNC caused an average lag of around 500\,ms.

@@

To describe the mission-critical performance aspects of the discussed system,
we initially provide a brief overview of potential challenges
by outlining bottlenecks and other relevant problem areas.
This is done in the context of the underlying use case model of DaaS systems.
Subsequently, we give a more comprehensive description
of our specific design decisions and tailored solutions.

Performance, from our viewpoint, stands out as the most crucial system property,
implying influence of other  relevant properties,
such as user experience, system responsiveness, and overall efficiency.
Consequently, we encountered multifaceted challenges
that potentially intersected across several layers within our architecture.

Hence, every component from the core infrastructure
to all backend services and user-facing elements,
must meet individual and domain-specific requirements
to contribute to the overall performance goals of the entire system.
Thus, we outline key limiting factors that hinder performance,
along with our experiences and the chosen mitigation strategies.

Given the absence of a generalized measurement system,
we ranked these factors based on their magnitude of impact,
drawing from our experiences obtained during the project.

% From our point of view, the three most critical performance aspects
% for Daas-Systems are:
% \begin{itemize}
% 	\item Infrastructure and Platform
% 	\item Network Latency and Bandwidth
% 	\item Internal Processing and Communication
% \end{itemize}
% Since each of them has their own characteristics,
% specific behavior and requirements, implying a potential impact
% on the overall system performance and resource utilization,
% we describe them individually and in more detail
% in the following separate subsections.

\subsubsection{Infrastructure and Platform}
In general can be said that the chosen infrastructure, platform,
and utilized software combined may have the most impact on the overall resource consumption
and are therefore of major importance.
As a DaaS system combines all aspects of IaaS-, PaaS, and SaaS-based computation
each individual performance characteristics has to be taken into account.

Regarding Infrastructure-as-a-Service (IaaS),
our solution generally offers the flexibility
to use either virtual machines or containers.
Scientific publications discussing relevant performance aspects of such systems
highlight the apparent advantage of a lightweight container solution
over virtual machines when identical programs are tested in both environments
~\cite{felter2015updated,potdar2020performance,seo2014performance}.

This suggests that a container-based approach is likely preferable
if the choice is available.
However, not every application can be run
in a container-based environment,
so we maximize coverage of potential usecases
by offering both container-based and virtual machine-based application hosting.

Regarding Platform-as-a-Service (PaaS),
studies show that there is no dramatic difference in performance
when derivatives of operating systems are compared to each other
~\cite{balen2020performance,boras2020performance}.
Nevertheless, some derivatives are more lightweight and performant.
In contrast, comparing operating systems of different types
reveals more obvious results, particularly when considering
low-level operating system functionality implementations
such as interrupt handling, memory allocation, and boot time
~\cite{sergeev2022docker,sulaiman2021comparison}.
Additional performance potentials can be leveraged
in cases where a Windows virtual machine can be replaced
with a container-based solution integrating wine
~\cite{huang2012performance}.  % win inside vm

For SaaS, performance largely depends on the underlying usage model
and the actual software being virtualized.
Depending on the use case, influential properties
such as multi-tenancy, provision strategies or the virtualized data type
might be utilized to varying degrees, resulting in overall
better or worse performing systems ~\cite{zhong2010virtualization}.

Despite the potential for usecase-specific optimization,
we emphasize that for DaaS systems,
coverage of a broad range of potential use cases is generally more favorable
than solving a particular use case in the most efficient way.
Therefore, we consider all user-facing desktop applications
as plausible candidates for migration to our systems,
rather than more complex distributed systems
for which a tailored environments might be more suitable.

As a result of these considerations, we aim to support
any Windows or Linux-based application in virtual machines
and support any Linux- and Wine-based applications for container environments.

\subsubsection{Network Latency and Bandwidth}
Network Latency and Bandwidth are crucial factors
in delivering web-based access to DaaS apps,
directly impacting system responsiveness experienced by users.
Supporting standardized frame-buffer protocols enables user interaction initially
and provide additional access to external devices like printers or USB sticks.
In cases involving for example video content transmission,
network bandwidth often becomes the main bottleneck,
especially with high-resolution video streams.
In contrast to that,  when transmitting audio, mouse movements,
or keyboard inputs, network latency becomes more important.

Application-specific requirements can be managed to some extent
through protocol-inherent tuning parameters and compression algorithms.
Our solution therefore offers a range of tuning options
in order to adapt to different use cases
and empowering users to choose their own tailored solution.

We utilize the Guacamole desktop gateway implementation,
supporting widely-used frame-buffer protocols like VNC and RDP,
along with relevant tuning options.
Guacamole also enables audio support and integration with external devices
through a JavaScript-based web client.

However, integrating Guacamole into the system
requires consideration of additional factors not handled by the software.
Authentication mechanisms for example are particularly important in a DaaS context,
as users expect seamless authentication across the whole system
without multiple logins per session.
Additionally, safeguarding sensitive information like user passwords is crucial
for maintaining system security and user trust.
To mitigate the risk of unintended system manipulation,
Guacamole was integrated with a proxied approach and by using websockets.
This ensures that credentials of a generated instance remain secret to the host system
while still providing authorized, passwordless access to the hosted application.

Moreover, the proxy mechanism enables support for other additional features
such as dynamic desktop resizing.
While Guacamole offers support for that by default in RDP,
support is completely missing for the VNC protocol.
Nevertheless, the proxied approach enables to intercept all exchanged websocket opcodes
and by listening for size requests
the requested feature could be added
by directly interacting with the instance on another layer.

While these changes to the default Guacamole use case are necessary to align
with the underlying DaaS design requirements,
they also negatively impact overall system performance and responsiveness.
The integration as a proxy introduces additional network hops,
leading to noticeable delays in network communication.
Additionally, the protocol-level evaluation of opcodes further increases this delay,
potentially affecting perceived responsiveness
if resize requests take longer than expected.
However, despite the introduced overhead,
the solution still enables platform-independent
and seamless access for all defined DaaS use cases.

\subsubsection{Internal processing and Communication}
Another critical aspect directly related to performance and responsiveness
is how data is internally processed, stored, and communicated
among all involved components and instances.
This includes metadata collected and processed by essential components
and involves direct exchange of raw data with object instances.

Ensuring a satisfying user experience in a DaaS system
implies specific components in a host system.
This includes access to all backend services
used to create containers and virtual machines,
local or remote databases, authentication services, filesystems, and network devices.
Additionally, a web server is integrated to provide relevant API endpoints
for frontend communication and websocket endpoints for viewer communication.

Generally spoken, all of these components can potentially become bottlenecks
when handling high throughputs or large amounts of concurrent applications
requiring low latency access and especially if all necessary backend components
are served from the same hardware.
Architecturally, all relevant components
must therefore be designed for distribution across different hosting systems.
However, retaining certain components directly on the host system
may still be required to ensure low-latency access.

From our observations, components requiring realtime and low-latency performance,
should physically be kept as close to the core hosting system as possible.
This includes all data and communication relevant to
audio, video, mouse, and keyboard transmission,
as well as all backend services for hosting containers and virtual machines.
Ideally, these components should be hosted natively on the host system
for virtual network and hardware access
or at least be connected to the same physical network
in order to minimize network latencies.

Regarding other components such as database access,
authentication providers, and filesystems,
maintaining a close distance from the host is advisable,
but this depends largely on the total number of users
as well as their specific use cases.
On smaller scales, keeping these components
in close physical proximity to the host remains prudent.

\subsection{Stability}
% TODO: Johannes (1)
% TODO: Christian (2)
% TODO: Johannes: Beschreibung zur Position der Komponenten, Möglichkeiten der Parallelisierung und Einschränkungen
%
% Was ist hier relevant?
% Verfügbarkeit des Service und der Benutzerdaten
% Redundante Komponenten (Hardware und Software)
% Welche Komponenten kann man sinnvoll redundant vorhalten?
% Was sind die Vor- und Nachteile?
% Auch Kosten?
% Wo sehen wir aktuell in unserer Architektur Risiken für Dienstausfall und Datenverlust?
% Features von Proxmox
% Ceph-Dateisystem
% Datenbank SQlite aktuell. Perspektivisch MariaDB
% Guacamole läuft 1x direkt auf dem Host aktuell. Läuft theoretisch auf jedem Host. Proxmox kann sich darum kümmern?!
% Docker haben/nicht haben / Vor- und Nachteile
% Hier auch auf die Skalierbarkeit beziehen
% Structure:
%	- Stability
%		- Redundancy
%		- Scalability
%	- Backend-services
%		- Database
%		- Authentication
%		- Distributed Filesystem
%		- Container/VM (Proxmox, Kubernetes?) 
%		- Instance error => Recreate
%	- Considerations:
%		- TODO: Mention that term: Divide and conquer
%		- Requirement: Distributable components
%	- Risks:
%		- Filesystem error
%		- Database error
%		- Authentication error
%		- Infrastructure error (Container/VM)
%	- Benefits/Drawbacks:
%		- Distribution of components is distribution of load => Performance
%		- Eliminates bottlenecks
%		- Errors affect only errornous component

Stability in DaaS applications is especially important
for meeting the demands of continuous and time-critical operations
across diverse user scenarios.
Ensuring the stability of all system components during runtime is paramount
to maintain service continuity, handle growing user loads,
and transparently recover from failures
while providing a seamless end-user experience.
These considerations are integral to our architectural concept,
directly impacting the operational integrity of the system
and its suitability for enterprise-level deployment and trust.

This definition of stability is supported
by various architectural and operational strategies,
including component redundancy, node scalability,
and a robust backend service framework facilitating communication
between components and nodes.
These strategies, along with their drawbacks and benefits,
will be further explored at a component level
and within the following subsections.

@@

Scalability in DaaS environments
can be approached through two fundamental methods: horizontal and vertical scaling.
Horizontal scaling, or scaling out, involves adding more machines or instances
to manage increased loads.
In contrast, vertical scaling, or scaling up,
adds more power (CPU, RAM) to existing machines
~\cite{vaquero2011dynamically}.
While horizontal scaling offers flexibility
and is often preferred in cloud environments,
vertical scaling remains important for applications
requiring strong single-thread performance or low latencies.

To comply with both approaches,
we designed the system and its backend services
to be split into logical units that can be distributed across dedicated systems.
Each backend service receives dedicated configuration
and can be accessed via hostname and port tuples.
By default, all backend services reside on the same host,
with tuples derived from the host configuration.
In a distributed context, services can be distributed to dedicated remote systems
to adhere to horizontal scaling principles.
Relevant tuples can then be derived from any node
using a common database containing meta information.

Vertical scaling of DaaS nodes is not currently planned,
but administrators are free to use custom third-party toolchains,
such as GPU virtualization, for vertical scaling on a per-host basis.
Although vertical scaling within a host is beyond the scope of our research project,
scaling the instances hosted within the system is feasible for both dimensions.

For vertical scaling of instances, we provide appropriate parameters
and API endpoints to manipulate resource assignments,
allowing for arbitrary configurations.
For horizontal scaling of instances, the built-in mechanisms
of each backend service are automatically configured
and utilized to distribute instances across a pool of connected nodes.
For virtual machines, Proxmox implements such mechanisms
as part of its 'Proxmox VE High Availability (HA)' cluster.
For containers, similar concepts can be integrated with any backend service,
such as Kubernetes.
Within these clusters,
DaaS instances are entirely agnostic to their hosting system
and only need to be reachable via a known hostname or IP as a single requirement.

@@

Redundancy is crucial in distributed systems,
particularly in DaaS architectures,
to safeguard against service interruptions and data loss
by duplicating essential components, functions, and data.
This design principle ensures high availability and reliability,
even during component failures.
In the following, we provide a list of components eligible for redundancy,
briefly describing the most relevant strategies in our architecture.

\begin{itemize}
	\item \textbf{Frontend/Backend:}
	      In our final version,
	      redundancy of the frontend and backend components
	      will enable configuration for distributing multiple instances
	      to dedicated systems and utilizing standardized load-balancing strategies.
	      This setup allows the frontend or backend
	      to serve as a gateway between the user and the backend service
	      containing all relevant instances.
	      Both components contribute to redundancy
	      by distributing requests and responses to their respecitive recipients
	      instead of processing them on their own.
	\item \textbf{Database:}
	      The database is integrated as a backend service
	      and is therefore independent of surrounding components and contexts.
	      It can be distributed to a dedicated system if necessary.
	      Modern database solutions, like MariaDB,
	      offer rich features for maintaining specific data consistency strategies
	      such as data replication, clustering, failover mechanisms,
	      and RAID control.
	      Currently, redundancies in our database system are not planned
	      but can be easily applied by users using such database-specific features
	      and a modified database adapter.
	\item \textbf{Authentication: }
	      Authentication within our architecture is implemented as a backend service,
	      allowing for distribution to any dedicated system.
	      It can also be targeted by standardized load-balancing mechanisms
	      and remains fully agnostic of surrounding components
	      or calling contexts.
	      The authentication system maintains its credential registry
	      in a separate database, along with a customizable set
	      of user and group privileges.
	      Additionally, it implements the well-defined OAuth 2.0 standard
	      to facilitate remote authentication.
	      In our final version,
	      a customizable amount of redundant systems may be configured.
	\item \textbf{Distributed Filesystem:}
	      Within our architecture, we utilize the distributed filesystem Ceph,
	      which offers object, block, and file storage capabilities.
	      Redundancy in Ceph is implicitly achieved
	      through its self-healing and self-managing features, employing
	      the CRUSH algorithm (Controlled Replication Under Scalable Hashing).
	      Data availability and fault tolerance are ensured
	      by automatically replicating data across different devices.
	      Our architecture leverages these capabilities
	      to reliably provide file storage to host systems
	      and all hosted DaaS instances,
	      while also offering block storage to the infrastructural backend services.
	      Hosting DaaS instances directly from the distributed block storage
	      enables migration and redundancy of instances or in other words, horizontal scaling.
	\item \textbf{Instances:} Redundancy of instances
	      It is not a necessary condition for DaaS systems in general,
	      but applying such strategies under certain conditions might be plausible.
	      By using highly available clusters within our backend services
	      redundancy can be achieved by using the corresponding API endpoints
	      previously described and by parameterizing them appropriately.
\end{itemize}

\subsubsection{Internal Communication}
After discussing the principles of scalability and redundancy,
we finally discuss how communication flows are managed in our distributed environment.
Our proposed solution is rooted in a message-oriented architecture
designed to fulfill DaaS properties for scalability and redundancy,
as previously outlined.

This implies that any action triggered, whether in the frontend on the client side,
backend on the host side, or within hosted instances,
must ensure that the necessary information for execution
is readily available at the time of processing.

We achieve this by storing runtime information in distributed block storage,
relevant configuration files in distributed file storage,
object metadata in a common database,
and by utilizing a decentralized OAuth component for authentication.

This ensures that any change triggered by a node within the DaaS domain
is immediately available to any other node relying in some way
on that particular information change.
Additionally, any extra information needed by a component must be fully contained
within the related request or response.
This enables arbitrarily distributed configurations as long as request
and response messages reach their desired target eventually
and as long relevant backend services remain available.

If that is the case, each distributed DaaS node might only contain essential and time-critical components,
including the backend service hosting relevant instances as well as the described
websocket proxy connected to the guacamole daemon and the user.
Any other component may reside elsewhere within the DaaS network.
